New cardinality estimation algorithms for HyperLogLog sketches

### 2016-12-18 paper/paper.tex
get rid of svg package, replaced png figures by pdf figures
some minor improvements

\documentclass[a4paper]{scrartcl}
\usepackage[utf8]{inputenc}
\usepackage{amsmath}
\usepackage{amssymb}
\usepackage{graphicx}
\usepackage{svg}
\usepackage{algpseudocode,algorithm,algorithmicx}
\usepackage{float}
\usepackage{url}
\usepackage{booktabs}
\usepackage{array}
\usepackage{paralist}
\usepackage{verbatim}
\usepackage{subfig}
\usepackage{hyperref}
\usepackage[binary-units = true]{siunitx}
\usepackage{csvsimple}
\usepackage{multirow}
\usepackage[capitalise,nameinlink,noabbrev]{cleveref}

@@

\begin{figure}
\centering
\includesvg[width=1\textwidth]{raw_estimate}
\caption{The distribution of the relative estimation error over the cardinality for the raw estimator after evaluation of \num{10000} randomly generated HyperLogLog data structures with parameters $\symPrecision = 12$ and $\symRegRange=20$.}
\label{fig:raw_estimate}
\end{figure}

\begin{figure}
\centering
\includesvg[width=1\textwidth]{small_range_estimate}
\caption{The distribution of the relative estimation error for the linear counting estimator after evaluation of \num{10000} randomly generated bitmaps of size $2^{12}$ which correspond to HyperLogLog sketches with parameters $\symPrecision = 12$ and $\symRegRange=0$.}
\label{fig:small_range_estimate}
\end{figure}

\begin{figure}
\centering
\includesvg[width=1\textwidth]{original_estimate}
\caption{The distribution of the relative estimation error over the cardinality for the original estimation algorithm after evaluation of \num{10000} randomly generated HyperLogLog data structures with parameters $\symPrecision = 12$ and $\symRegRange=20$.}
\label{fig:original_estimate}
\end{figure}

@@

\begin{figure}
\centering
\includesvg[width=0.6\textwidth]{power-series-function-minus-1}
\caption{The deviation of $\symPowerSeriesFunc(\symX)$ from 1.}
\label{fig:power_series_func}
\end{figure}

@@

\begin{figure}
\centering
\includesvg[width=1\textwidth]{raw_corrected_estimate_12_20}
\caption{Relative error of the corrected raw estimator as a function of the true cardinality for a HyperLogLog sketch with parameters $\symPrecision = 12$ and $\symRegRange=20$.}
\label{fig:raw_corrected_estimation_error_12_20}
\end{figure}

@@

\begin{figure}
\centering
\includesvg[width=1\textwidth]{stdev_comparison}
\caption{Standard deviations of the relative error of different cardinality estimators over the true cardinality for a HyperLogLog sketch with paramters $\symPrecision = 12$ and $\symRegRange=20$ .}
\label{fig:stdev_comparison}
\end{figure}

@@

\begin{figure}
\centering
\includesvg[width=1\textwidth]{raw_corrected_estimate_8_24}
\caption{Relative error of the corrected raw estimator as a function of the true cardinality for a HyperLogLog sketch with parameters $\symPrecision = 8$ and $\symRegRange=24$.}
\label{fig:raw_corrected_estimation_error_8_24}
\end{figure}

\begin{figure}
\centering
\includesvg[width=1\textwidth]{raw_corrected_estimate_16_16}
\caption{Relative error of the corrected raw estimator as a function of the true cardinality for a HyperLogLog sketch with parameters $\symPrecision = 16$ and $\symRegRange=16$.}
\label{fig:raw_corrected_estimation_error_16_16}
\end{figure}

\begin{figure}
\centering
\includesvg[width=1\textwidth]{raw_corrected_estimate_22_10}
\caption{Relative error of the corrected raw estimator as a function of the true cardinality for a HyperLogLog sketch with parameters $\symPrecision = 22$ and $\symRegRange=10$.}
\label{fig:raw_corrected_estimation_error_22_10}
\end{figure}

@@

\begin{figure}
\centering
\includesvg[width=1\textwidth]{raw_corrected_estimate_12_52}
\caption{Relative error of the corrected raw estimator as a function of the true cardinality for a HyperLogLog sketch with parameters $\symPrecision = 12$ and $\symRegRange=52$.}
\label{fig:raw_corrected_estimation_error_12_52}
\end{figure}

@@

\begin{figure}
\centering
\includesvg[width=1\textwidth]{raw_corrected_estimate_12_14}
\caption{Relative error of the corrected raw estimator as a function of the true cardinality for a HyperLogLog sketch with parameters $\symPrecision = 12$ and $\symRegRange=14$.}
\label{fig:raw_corrected_estimation_error_12_14}
\end{figure}

@@

\begin{figure}
\centering
\includesvg[width=1\textwidth]{corrected_raw_avg_exec_time}
\caption{Average computation time as a function of the true cardinality with an Intel Core i5-2500K clocking at \SI{3.3}{\giga\hertz} when estimating the cardinality from HyperLogLog sketches with parameters $\symPrecision=12$, $\symRegRange=20$ and $\symPrecision=12$, $\symRegRange=52$, respectively. Both cases, $\symSmallCorrectionFunc$ and $\symLargeCorrectionFunc$ precalculated and calculated on-demand have been considered.}
\label{fig:corrected_raw_avg_exec_time}
\end{figure}

@@

We remark that we are not the first applying the maximum likelihood method to HyperLogLog sketches \cite{Clifford2012}, where the case without stochastic averaging (compare \cref{sec:data_element_insertion}) was considered. The register values are statistically independent by nature in this case which 
allows factorization of the joint probability mass function and allows a straightforward application of the maximum likelihood method.

The more relevant case with stochastic averaging, which is considered in this paper, is more complicated. However, with the help of the Poisson approximation, it is still feasible to apply the maximum likelihood method and we are able to derive a new robust and efficient cardinality estimation algorithm. Furthermore, we will demonstrate that consequent application of the maximum likelihood method reveals that the cardinality estimate needs to be roughly proportional to the harmonic mean for intermediate cardinality values. The history of the HyperLogLog algorithm shows that the raw estimator \eqref{equ:raw_estimator} was first found after several attempts using the geometric mean \cite{Flajolet2007, Durand2003}.

@@

Using the probability mass function of the Poisson model \eqref{equ:poisson_pmf} the log-likelihood and its derivative are given by
\begin{equation}
\label{equ:log_likelihood_single}
\log \mathcal{\symLikelihood}(\symPoissonRate\vert\boldsymbol{\symRegValVariate}) = 
-\frac{\symPoissonRate}{\symNumReg}\sum_{\symRegVal=0}^{\symRegRange}\frac{\symCountVariate_\symRegVal}{2^\symRegVal}+ 
\sum_{\symRegVal=1}^{\symRegRange}\symCountVariate_\symRegVal \log\!\left(1-e^{-\frac{\symPoissonRate}{\symNumReg 2^\symRegVal}}\right)
+
\symCountVariate_{\symRegRange+1} \log\!\left(1-e^{-\frac{\symPoissonRate}{\symNumReg 2^{\symRegRange}}}\right)
\end{equation}
and
\begin{equation}
\frac{d}{d\symPoissonRate}\log \mathcal{\symLikelihood}(\symPoissonRate\vert\boldsymbol{\symRegValVariate}) 
=
-\frac{1}{\symPoissonRate}\left(
\frac{\symPoissonRate}{\symNumReg}\sum_{\symRegVal=0}^\symRegRange \frac{\symCountVariate_\symRegVal}{2^\symRegVal}+
\sum_{\symRegVal=1}^\symRegRange \symCountVariate_\symRegVal\frac{\frac{\symPoissonRate}{\symNumReg 2^\symRegVal}}{1-e^{\frac{\symPoissonRate}{\symNumReg 2^\symRegVal}}}
+\symCountVariate_{\symRegRange+1}\frac{\frac{\symPoissonRate}{\symNumReg 2^\symRegRange}}{1-e^{\frac{\symPoissonRate}{\symNumReg 2^\symRegRange}}}
\right).
\end{equation}
As a consequence, the maximum likelihood estimate for the Poisson parameter is given by 
\begin{equation}
\symPoissonRateEstimate = \symNumReg\symXEstimate,
\end{equation}
if $\symXEstimate$ denotes the root of the function
\begin{equation}
\label{equ:funcdef}
\symFunc(\symX)
:=
\symX\sum_{\symRegVal=0}^\symRegRange \frac{\symCountVariate_\symRegVal}{2^\symRegVal}+
\sum_{\symRegVal=1}^\symRegRange \symCountVariate_\symRegVal\frac{\frac{\symX}{2^\symRegVal}}{1-e^{\frac{\symX}{2^\symRegVal}}}
+\symCountVariate_{\symRegRange+1}\frac{\frac{\symX}{2^\symRegRange}}{1-e^{\frac{\symX}{2^\symRegRange}}}.
\end{equation}
This function can also be written as
\begin{equation}
\label{equ:func}
\symFunc(\symX)
:=
\symX\sum_{\symRegVal=0}^\symRegRange \frac{\symCountVariate_\symRegVal}{2^\symRegVal}+
\sum_{\symRegVal=1}^\symRegRange \symCountVariate_\symRegVal\symHelper\!\left(\frac{\symX}{2^\symRegVal}\right)
+
\symCountVariate_{\symRegRange+1}\symHelper\!\left(\frac{\symX}{2^\symRegRange}\right)
-
\left(\symNumReg-\symCountVariate_0\right),
\end{equation}
where the function $\symHelper(\symX)$ is defined as
\begin{equation}
\label{equ:helper}
\symHelper(\symX):=1-\frac{\symX}{e^{\symX}-1}.
\end{equation}
$\symHelper(\symX)$ is strictly increasing and concave as can be seen in \cref{fig:helper_function}. For non-negative values $\symX$ the function ranges from $\symHelper(0)=0$ to $\symHelper(\symX\rightarrow \infty)=1$.
\begin{figure}
\centering
\includesvg[width=0.6\textwidth]{helper}
\caption{The function $\symHelper(\symX)$.}
\label{fig:helper_function}
\end{figure}
Since the function $\symFunc(x)$ is also strictly increasing, it is obvious that there exists a unique root $\symXEstimate$ for which $\symFunc(\symXEstimate)=0$. The function is non-positive at 0 since $\symFunc(0)=\symCountVariate_0-\symNumReg\leq 0$ and, in case $\symCountVariate_{\symRegRange+1}<\symNumReg$ which implies $\sum_{\symRegVal=0}^\symRegRange \frac{\symCountVariate_\symRegVal}{2^\symRegVal}>0$, the function is at least linearly increasing. $\symCountVariate_{\symRegRange+1}=\symNumReg$ corresponds to the case with all registers equal to the maximum value $\symRegRange+1$, for which the maximum likelihood estimate would be positive infinite.

@@

\subsection{Computation of the maximum likelihood estimate}
\label{sec:comp_ml_estimate}
Since $\symFunc$ is concave and increasing, both, Newton-Raphson iteration and the secant method, will converge to the root, provided that the function is negative for the starting points. In the following we start from the secant method to derive the new cardinality estimation algorithm. Even though the secant method has the disadvantage of slower convergence, a single iteration is simpler to calculate as it does not require the evaluation of the first derivative. An iteration step of the secant method can be written as
\begin{equation}
\symX_{\symIndexI} = 
\symX_{\symIndexI-1} -
\left(\symX_{\symIndexI-1}-\symX_{\symIndexI-2}\right)
\frac{\symFunc(\symX_{\symIndexI-1})}{\symFunc(\symX_{\symIndexI-1}) - \symFunc(\symX_{\symIndexI-2})}.
\end{equation}
If we set $\symX_0$ equal to 0, for which  $\symFunc(\symX_0)=-\left(\symNumReg-\symCountVariate_0\right)$, and $\symX_1$ equal to one of the derived lower bounds \eqref{equ:strong_lower_bound} or \eqref{equ:weak_lower_bound}, the sequence $(\symX_0, \symX_1, \symX_2, \ldots)$ is montone increasing. Using the definitions
\begin{equation}
\Delta\symX_\symIndexI := \symX_\symIndexI-\symX_{\symIndexI-1}
\end{equation}
and
\begin{equation}
\label{equ:funcprime}
\symFuncPrime(\symX):=\symFunc(\symX) + \left(\symNumReg-\symCountVariate_0\right)
=\symX\sum_{\symRegVal=0}^\symRegRange \frac{\symCountVariate_\symRegVal}{2^\symRegVal}+
\sum_{\symRegVal=1}^\symRegRange \symCountVariate_\symRegVal\symHelper\!\left(\frac{\symX}{2^\symRegVal}\right)
+
\symCountVariate_{\symRegRange+1}\symHelper\!\left(\frac{\symX}{2^\symRegRange}\right)
\end{equation}
the iteration scheme can also be written as
\begin{gather}
\label{equ:secant_delta}
\Delta\symX_{\symIndexI} = \Delta\symX_{\symIndexI-1}
\frac{\left(\symNumReg-\symCountVariate_0\right)-\symFuncPrime(\symX_{\symIndexI-1})}{\symFuncPrime(\symX_{\symIndexI-1}) - \symFuncPrime(\symX_{\symIndexI-2})},
\\
\symX_{\symIndexI} = \symX_{\symIndexI-1} + \Delta\symX_{\symIndexI}.
\end{gather}
The iteration can be stopped, if $\Delta\symX_{\symIndexI} \leq \symStopDelta\cdot \symX_{\symIndexI}$. Since the expected statistical errror for the HyperLogLog data structure scales according to $\frac{1}{\sqrt{\symNumReg}}$ \cite{Flajolet2007}, it makes sense to choose $\symStopDelta = \frac{\symStopEpsilon}{\sqrt{\symNumReg}}$ with some constant $\symStopEpsilon$. For all results that will be presented in \cref{sec:maximum_likelihood_estimation_error} we have used $\symStopEpsilon = 10^{-2}$.

@@

Putting all these optimizations together finally gives the new cardinality estimation algorithm presented in \cref{alg:estimate_ml}. The algorithm requires mainly only elementary operations. The square root that appears for the calculation of the stopping limit $\symStopDelta=\symStopEpsilon/\sqrt{\symNumReg}$ can be precomputed, because it only depends on the HyperLogLog precision parameter $\symPrecision$. For large cardinalities it is worth to take the strong \eqref{equ:strong_lower_bound} instead of the weak lower bound \eqref{equ:weak_lower_bound} as initial value, which requires a single logarithm evaluation. The stronger bound is a much better approximation especially for large cardinalities, where the extra logarithm evaluation is amortized by savings in the number of iteration cycles.

@@

\begin{figure}
\centering
\includesvg[width=1\textwidth]{max_likelihood_estimate_12_20}
\caption{Relative error of the maximum likelihood estimates as a function of the true cardinality for a HyperLogLog sketch with parameters $\symPrecision = 12$ and $\symRegRange=20$.}
\label{fig:max_likelihood_estimation_error_12_20}
\end{figure}

\begin{figure}
\centering
\includesvg[width=1\textwidth]{max_likelihood_estimate_8_24}
\caption{Relative error of the maximum likelihood estimates as a function of the true cardinality for a HyperLogLog sketch with parameters $\symPrecision = 8$ and $\symRegRange=24$.}
\label{fig:max_likelihood_estimation_error_8_24}
\end{figure}

\begin{figure}
\centering
\includesvg[width=1\textwidth]{max_likelihood_estimate_16_16}
\caption{Relative error of the maximum likelihood estimates as a function of the true cardinality for a HyperLogLog sketch with parameters $\symPrecision = 16$ and $\symRegRange=16$.}
\label{fig:max_likelihood_estimation_error_16_16}
\end{figure}

\begin{figure}
\centering
\includesvg[width=1\textwidth]{max_likelihood_estimate_22_10}
\caption{Relative error of the maximum likelihood estimates as a function of the true cardinality for a HyperLogLog sketch with parameters $\symPrecision = 22$ and $\symRegRange=10$.}
\label{fig:max_likelihood_estimation_error_22_10}
\end{figure}

\begin{figure}
\centering
\includesvg[width=1\textwidth]{max_likelihood_estimate_12_52}
\caption{Relative error of the maximum likelihood estimates as a function of the true cardinality for a HyperLogLog sketch with parameters $\symPrecision = 12$ and $\symRegRange=52$.}
\label{fig:max_likelihood_estimation_error_12_52}
\end{figure}

\begin{figure}
\centering
\includesvg[width=1\textwidth]{max_likelihood_estimate_12_14}
\caption{Relative error of the maximum likelihood estimates as a function of the true cardinality for a HyperLogLog sketch with parameters $\symPrecision = 12$ and $\symRegRange=14$.}
\label{fig:max_likelihood_estimation_error_12_14}
\end{figure}

@@

\begin{figure}
\centering
\includesvg[width=1\textwidth]{max_likelihood_avg_exec_time}
\caption{Average execution time of the maximum likelihood estimation algorithm as a function of the true cardinality with an Intel Core i5-2500K clocking at \SI{3.3}{\giga\hertz} for HyperLogLog sketches with parameters $\symPrecision=12$, $\symRegRange=20$ and $\symPrecision=12$, $\symRegRange=52$, respectively.}
\label{fig:avg_exec_time}
\end{figure}

>>>>

\documentclass[a4paper]{scrartcl}
\usepackage[utf8]{inputenc}
\usepackage{amsmath}
\usepackage{amssymb}
\usepackage{graphicx}
\usepackage{algpseudocode,algorithm,algorithmicx}
\usepackage{float}
\usepackage{url}
\usepackage{booktabs}
\usepackage{array}
\usepackage{paralist}
\usepackage{verbatim}
\usepackage{subfig}
\usepackage{hyperref}
\usepackage[binary-units = true]{siunitx}
\usepackage{csvsimple}
\usepackage{multirow}
\usepackage[capitalise,nameinlink,noabbrev]{cleveref}

@@

\begin{figure}
\centering
\includegraphics[width=1\textwidth]{raw_estimate}
\caption{The distribution of the relative estimation error over the cardinality for the raw estimator after evaluation of \num{10000} randomly generated HyperLogLog data structures with parameters $\symPrecision = 12$ and $\symRegRange=20$.}
\label{fig:raw_estimate}
\end{figure}

\begin{figure}
\centering
\includegraphics[width=1\textwidth]{small_range_estimate}
\caption{The distribution of the relative estimation error for the linear counting estimator after evaluation of \num{10000} randomly generated bitmaps of size $2^{12}$ which correspond to HyperLogLog sketches with parameters $\symPrecision = 12$ and $\symRegRange=0$.}
\label{fig:small_range_estimate}
\end{figure}

\begin{figure}
\centering
\includegraphics[width=1\textwidth]{original_estimate}
\caption{The distribution of the relative estimation error over the cardinality for the original estimation algorithm after evaluation of \num{10000} randomly generated HyperLogLog data structures with parameters $\symPrecision = 12$ and $\symRegRange=20$.}
\label{fig:original_estimate}
\end{figure}

@@

\begin{figure}
\centering
\includegraphics[width=0.6\textwidth]{power-series-function-minus-1}
\caption{The deviation of $\symPowerSeriesFunc(\symX)$ from 1.}
\label{fig:power_series_func}
\end{figure}

@@

\begin{figure}
\centering
\includegraphics[width=1\textwidth]{raw_corrected_estimate_12_20}
\caption{Relative error of the corrected raw estimator as a function of the true cardinality for a HyperLogLog sketch with parameters $\symPrecision = 12$ and $\symRegRange=20$.}
\label{fig:raw_corrected_estimation_error_12_20}
\end{figure}

@@

\begin{figure}
\centering
\includegraphics[width=1\textwidth]{stdev_comparison}
\caption{Standard deviations of the relative error of different cardinality estimators over the true cardinality for a HyperLogLog sketch with paramters $\symPrecision = 12$ and $\symRegRange=20$ .}
\label{fig:stdev_comparison}
\end{figure}

@@

\begin{figure}
\centering
\includegraphics[width=1\textwidth]{raw_corrected_estimate_8_24}
\caption{Relative error of the corrected raw estimator as a function of the true cardinality for a HyperLogLog sketch with parameters $\symPrecision = 8$ and $\symRegRange=24$.}
\label{fig:raw_corrected_estimation_error_8_24}
\end{figure}

\begin{figure}
\centering
\includegraphics[width=1\textwidth]{raw_corrected_estimate_16_16}
\caption{Relative error of the corrected raw estimator as a function of the true cardinality for a HyperLogLog sketch with parameters $\symPrecision = 16$ and $\symRegRange=16$.}
\label{fig:raw_corrected_estimation_error_16_16}
\end{figure}

\begin{figure}
\centering
\includegraphics[width=1\textwidth]{raw_corrected_estimate_22_10}
\caption{Relative error of the corrected raw estimator as a function of the true cardinality for a HyperLogLog sketch with parameters $\symPrecision = 22$ and $\symRegRange=10$.}
\label{fig:raw_corrected_estimation_error_22_10}
\end{figure}

@@

\begin{figure}
\centering
\includegraphics[width=1\textwidth]{raw_corrected_estimate_12_52}
\caption{Relative error of the corrected raw estimator as a function of the true cardinality for a HyperLogLog sketch with parameters $\symPrecision = 12$ and $\symRegRange=52$.}
\label{fig:raw_corrected_estimation_error_12_52}
\end{figure}

@@

\begin{figure}
\centering
\includegraphics[width=1\textwidth]{raw_corrected_estimate_12_14}
\caption{Relative error of the corrected raw estimator as a function of the true cardinality for a HyperLogLog sketch with parameters $\symPrecision = 12$ and $\symRegRange=14$.}
\label{fig:raw_corrected_estimation_error_12_14}
\end{figure}

@@

\begin{figure}
\centering
\includegraphics[width=1\textwidth]{corrected_raw_avg_exec_time}
\caption{Average computation time as a function of the true cardinality with an Intel Core i5-2500K clocking at \SI{3.3}{\giga\hertz} when estimating the cardinality from HyperLogLog sketches with parameters $\symPrecision=12$, $\symRegRange=20$ and $\symPrecision=12$, $\symRegRange=52$, respectively. Both cases, $\symSmallCorrectionFunc$ and $\symLargeCorrectionFunc$ precalculated and calculated on-demand have been considered.}
\label{fig:corrected_raw_avg_exec_time}
\end{figure}

@@

For HyperLogLog sketches that have been obtained without stochastic averaging (see \cref{sec:data_element_insertion}) the maximum likelihood method was already previously considered for cardinality estimation \cite{Clifford2012}. However, in this case the register values are statistically independent by nature which 
allows factorization of the joint probability mass function and a straightforward calculation of the maximum likelihood estimate. The practically more relevant stochastic averaging case, which is considered in this paper, gives a more complicated likelihood function. However, the Poisson approximation makes the maximum likelihood method feasible again and we are finally able to derive another new robust and efficient cardinality estimation algorithm. Furthermore, in the course of the derivation we will demonstrate that consequent application of the maximum likelihood method reveals that the cardinality estimate needs to be roughly proportional to the harmonic mean for intermediate cardinality values. The history of the HyperLogLog algorithm shows that the raw estimator \eqref{equ:raw_estimator} was first found after several attempts using the geometric mean \cite{Flajolet2007, Durand2003}.

@@

Using the probability mass function of the Poisson model \eqref{equ:poisson_pmf} the log-likelihood and its derivative are given by
\begin{equation}
\label{equ:log_likelihood_single}
\log \mathcal{\symLikelihood}(\symPoissonRate\vert\boldsymbol{\symRegValVariate}) = 
-\frac{\symPoissonRate}{\symNumReg}\sum_{\symRegVal=0}^{\symRegRange}\frac{\symCountVariate_\symRegVal}{2^\symRegVal}+ 
\sum_{\symRegVal=1}^{\symRegRange}\symCountVariate_\symRegVal \log\!\left(1-e^{-\frac{\symPoissonRate}{\symNumReg 2^\symRegVal}}\right)
+
\symCountVariate_{\symRegRange+1} \log\!\left(1-e^{-\frac{\symPoissonRate}{\symNumReg 2^{\symRegRange}}}\right)
\end{equation}
and
\begin{equation}
\frac{d}{d\symPoissonRate}\log \mathcal{\symLikelihood}(\symPoissonRate\vert\boldsymbol{\symRegValVariate}) 
=
-\frac{1}{\symPoissonRate}\left(
\frac{\symPoissonRate}{\symNumReg}\sum_{\symRegVal=0}^\symRegRange \frac{\symCountVariate_\symRegVal}{2^\symRegVal}+
\sum_{\symRegVal=1}^\symRegRange \symCountVariate_\symRegVal\frac{\frac{\symPoissonRate}{\symNumReg 2^\symRegVal}}{1-e^{\frac{\symPoissonRate}{\symNumReg 2^\symRegVal}}}
+\symCountVariate_{\symRegRange+1}\frac{\frac{\symPoissonRate}{\symNumReg 2^\symRegRange}}{1-e^{\frac{\symPoissonRate}{\symNumReg 2^\symRegRange}}}
\right).
\end{equation}
As a consequence, the maximum likelihood estimate for the Poisson parameter is given by 
\begin{equation}
\symPoissonRateEstimate = \symNumReg\symXEstimate,
\end{equation}
if $\symXEstimate$ denotes the root of the function
\begin{equation}
\label{equ:funcdef}
\symFunc(\symX)
:=
\symX\sum_{\symRegVal=0}^\symRegRange \frac{\symCountVariate_\symRegVal}{2^\symRegVal}+
\sum_{\symRegVal=1}^\symRegRange \symCountVariate_\symRegVal\frac{\frac{\symX}{2^\symRegVal}}{1-e^{\frac{\symX}{2^\symRegVal}}}
+\symCountVariate_{\symRegRange+1}\frac{\frac{\symX}{2^\symRegRange}}{1-e^{\frac{\symX}{2^\symRegRange}}}.
\end{equation}
This function can also be written as
\begin{equation}
\label{equ:func}
\symFunc(\symX)
:=
\symX\sum_{\symRegVal=0}^\symRegRange \frac{\symCountVariate_\symRegVal}{2^\symRegVal}+
\sum_{\symRegVal=1}^\symRegRange \symCountVariate_\symRegVal\symHelper\!\left(\frac{\symX}{2^\symRegVal}\right)
+
\symCountVariate_{\symRegRange+1}\symHelper\!\left(\frac{\symX}{2^\symRegRange}\right)
-
\left(\symNumReg-\symCountVariate_0\right),
\end{equation}
where the function $\symHelper(\symX)$ is defined as
\begin{equation}
\label{equ:helper}
\symHelper(\symX):=1-\frac{\symX}{e^{\symX}-1}.
\end{equation}
$\symHelper(\symX)$ is strictly increasing and concave as can be seen in \cref{fig:helper_function}. For non-negative values $\symX$ the function ranges from $\symHelper(0)=0$ to $\symHelper(\symX\rightarrow \infty)=1$.
\begin{figure}
\centering
\includegraphics[width=0.6\textwidth]{helper}
\caption{The function $\symHelper(\symX)$.}
\label{fig:helper_function}
\end{figure}
Since the function $\symFunc(x)$ is also strictly increasing, it is obvious that there exists a unique root $\symXEstimate$ for which $\symFunc(\symXEstimate)=0$. The function is non-positive at 0 since $\symFunc(0)=\symCountVariate_0-\symNumReg\leq 0$ and, in case $\symCountVariate_{\symRegRange+1}<\symNumReg$ which implies $\sum_{\symRegVal=0}^\symRegRange \frac{\symCountVariate_\symRegVal}{2^\symRegVal}>0$, the function is at least linearly increasing. $\symCountVariate_{\symRegRange+1}=\symNumReg$ corresponds to the case with all registers equal to the maximum value $\symRegRange+1$, for which the maximum likelihood estimate would be positive infinite.

@@

\subsection{Computation of the maximum likelihood estimate}
\label{sec:comp_ml_estimate}
Since $\symFunc$ is concave and increasing, both, Newton-Raphson iteration and the secant method, will converge to the root, provided that the function is negative for the starting points. In the following we start from the secant method to derive the new cardinality estimation algorithm. Even though the secant method has the disadvantage of slower convergence, a single iteration is simpler to calculate as it does not require the evaluation of the first derivative. An iteration step of the secant method can be written as
\begin{equation}
\symX_{\symIndexI} = 
\symX_{\symIndexI-1} -
\left(\symX_{\symIndexI-1}-\symX_{\symIndexI-2}\right)
\frac{\symFunc(\symX_{\symIndexI-1})}{\symFunc(\symX_{\symIndexI-1}) - \symFunc(\symX_{\symIndexI-2})}.
\end{equation}
If we set $\symX_0$ equal to 0, for which  $\symFunc(\symX_0)=-\left(\symNumReg-\symCountVariate_0\right)$, and $\symX_1$ equal to one of the derived lower bounds \eqref{equ:strong_lower_bound} or \eqref{equ:weak_lower_bound}, the sequence $(\symX_0, \symX_1, \symX_2, \ldots)$ is montone increasing. Using the definitions
\begin{equation}
\Delta\symX_\symIndexI := \symX_\symIndexI-\symX_{\symIndexI-1}
\end{equation}
and
\begin{equation}
\label{equ:funcprime}
\symFuncPrime(\symX):=\symFunc(\symX) + \left(\symNumReg-\symCountVariate_0\right)
=\symX\sum_{\symRegVal=0}^\symRegRange \frac{\symCountVariate_\symRegVal}{2^\symRegVal}+
\sum_{\symRegVal=1}^\symRegRange \symCountVariate_\symRegVal\symHelper\!\left(\frac{\symX}{2^\symRegVal}\right)
+
\symCountVariate_{\symRegRange+1}\symHelper\!\left(\frac{\symX}{2^\symRegRange}\right)
\end{equation}
the iteration scheme can also be written as
\begin{gather}
\label{equ:secant_delta}
\Delta\symX_{\symIndexI} = \Delta\symX_{\symIndexI-1}
\frac{\left(\symNumReg-\symCountVariate_0\right)-\symFuncPrime(\symX_{\symIndexI-1})}{\symFuncPrime(\symX_{\symIndexI-1}) - \symFuncPrime(\symX_{\symIndexI-2})},
\\
\symX_{\symIndexI} = \symX_{\symIndexI-1} + \Delta\symX_{\symIndexI}.
\end{gather}
The iteration can be stopped, if $\Delta\symX_{\symIndexI} \leq \symStopDelta\cdot \symX_{\symIndexI}$. Since the expected statistical errror for the HyperLogLog data structure scales according to $\frac{1}{\sqrt{\symNumReg}}$ \cite{Flajolet2007}, it makes sense to choose $\symStopDelta = \frac{\symStopEpsilon}{\sqrt{\symNumReg}}$ with some constant $\symStopEpsilon$. For all results presented later in \cref{sec:maximum_likelihood_estimation_error} we have used $\symStopEpsilon = 10^{-2}$.

@@

Putting all these optimizations together finally gives the new cardinality estimation algorithm presented as \cref{alg:estimate_ml}. The algorithm requires mainly only elementary operations. The square root that appears for the calculation of the stopping limit $\symStopDelta=\symStopEpsilon/\sqrt{\symNumReg}$ can be precomputed, because it only depends on the HyperLogLog precision parameter $\symPrecision$. For large cardinalities it is worth to take the strong \eqref{equ:strong_lower_bound} instead of the weak lower bound \eqref{equ:weak_lower_bound} as initial value, which requires a single logarithm evaluation. The stronger bound is a much better approximation especially for large cardinalities, where the extra logarithm evaluation is amortized by savings in the number of iteration cycles.

@@

\begin{figure}
\centering
\includegraphics[width=1\textwidth]{max_likelihood_estimate_12_20}
\caption{Relative error of the maximum likelihood estimates as a function of the true cardinality for a HyperLogLog sketch with parameters $\symPrecision = 12$ and $\symRegRange=20$.}
\label{fig:max_likelihood_estimation_error_12_20}
\end{figure}

\begin{figure}
\centering
\includegraphics[width=1\textwidth]{max_likelihood_estimate_8_24}
\caption{Relative error of the maximum likelihood estimates as a function of the true cardinality for a HyperLogLog sketch with parameters $\symPrecision = 8$ and $\symRegRange=24$.}
\label{fig:max_likelihood_estimation_error_8_24}
\end{figure}

\begin{figure}
\centering
\includegraphics[width=1\textwidth]{max_likelihood_estimate_16_16}
\caption{Relative error of the maximum likelihood estimates as a function of the true cardinality for a HyperLogLog sketch with parameters $\symPrecision = 16$ and $\symRegRange=16$.}
\label{fig:max_likelihood_estimation_error_16_16}
\end{figure}

\begin{figure}
\centering
\includegraphics[width=1\textwidth]{max_likelihood_estimate_22_10}
\caption{Relative error of the maximum likelihood estimates as a function of the true cardinality for a HyperLogLog sketch with parameters $\symPrecision = 22$ and $\symRegRange=10$.}
\label{fig:max_likelihood_estimation_error_22_10}
\end{figure}

\begin{figure}
\centering
\includegraphics[width=1\textwidth]{max_likelihood_estimate_12_52}
\caption{Relative error of the maximum likelihood estimates as a function of the true cardinality for a HyperLogLog sketch with parameters $\symPrecision = 12$ and $\symRegRange=52$.}
\label{fig:max_likelihood_estimation_error_12_52}
\end{figure}

\begin{figure}
\centering
\includegraphics[width=1\textwidth]{max_likelihood_estimate_12_14}
\caption{Relative error of the maximum likelihood estimates as a function of the true cardinality for a HyperLogLog sketch with parameters $\symPrecision = 12$ and $\symRegRange=14$.}
\label{fig:max_likelihood_estimation_error_12_14}
\end{figure}

@@

\begin{figure}
\centering
\includegraphics[width=1\textwidth]{max_likelihood_avg_exec_time}
\caption{Average execution time of the maximum likelihood estimation algorithm as a function of the true cardinality with an Intel Core i5-2500K clocking at \SI{3.3}{\giga\hertz} for HyperLogLog sketches with parameters $\symPrecision=12$, $\symRegRange=20$ and $\symPrecision=12$, $\symRegRange=52$, respectively.}
\label{fig:avg_exec_time}
\end{figure}
